---
title: "Sparse Multiple Index (SMI) Models for High-dimensional Nonparametric Forecasting"
author: Nuwani Palihawadana
date: 01 July 2024
titlegraphic: bg-13.png
titlecolor: white
toc: true
format:
  presentation-beamer:
    pdf-engine: xelatex
    template-partials:
        - before-title.tex
execute:
  cache: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| echo: false
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tsibble)
library(lubridate)
library(cowplot)
```

# Motivation

## Electricity Demand Data
```{r, echo=FALSE, message=FALSE}
#| echo: false
#| message: false

demand_data <- tsibbledata::vic_elec %>%
  mutate(
    Half_hour = hms(format(Time, "%H:%M:%S"))
  ) %>%
  as_tsibble(index = Time, key = Half_hour)

p1 <- demand_data %>%
  ggplot(aes(x=Time, y=Demand)) +
  geom_line(color = "#D55E00") +
  xlab("Time")+
  ylab("Demand (MW)") +
  ggtitle("Half-hourly Electricity Demand - Victoria") +
  theme(
    plot.title = element_text(size = 10),
    ) +
  scale_x_datetime(date_breaks = "1 year") +
  ylim(3000, 8000)

dd <- tsibbledata::vic_elec %>%
  filter(Time <= "2012-01-31")

p2 <- dd %>%
  ggplot(aes(x=Time, y=Demand)) +
  geom_line(color = "#D55E00") +
  xlab("Date in January")+
  ylab("Demand (MW)") +
  ggtitle("Half-hourly Electricity Demand - Victoria - January 2012") +
  theme(
    plot.title = element_text(size = 10),
    ) +
  scale_x_datetime(date_breaks = "2 days", date_labels = "%d") +
  ylim(3000, 8000)

plot_grid(p1, p2, nrow=1)
```


## Electricity Demand Data
\placefig{1.8}{1.2}{width=14cm, height=7cm}{figs/elec_temp.pdf}

## Background
\begin{itemize}
  \item \color{violet} \textbf{\textsl{Nonlinear "Transfer Function" model}}
\end{itemize}
<!-- \begin{block}{} -->
<!--   \centerline{$y_{t} = f(\bm{x}_{t}, \bm{x}_{t-1}, \dots,\bm{x}_{t-p}, y_{1},\dots,y_{t-k}) + \varepsilon_{t}$} -->
<!-- \end{block} -->

$y_{t}$ -- variable to forecast

$\mathbfit{x}_{t}$ -- a vector of predictors

$\varepsilon_{t}$ -- random error

\pause
When we consider
\begin{block}{}
  \centerline{$f(\mathbfit{x}_{t}, \mathbfit{x}_{t-1}, \dots,\mathbfit{x}_{t-p})$}
\end{block}
\pause
\begin{itemize}
  \item Impossible to estimate $f$ for large $p$ -- \textbf{\textit{curse of dimensionality}}
  \pause
  \item Reasonable to impose additivity constraints \\[0.2cm]
\end{itemize}
<!-- \begin{block}{} -->
<!--   \centerline{$f(\bm{x}_{t}, \bm{x}_{t-1}, \dots,\bm{x}_{t-p}) = \sum_{i=0}^{p} {f_{i}(\bm{x}_{t-i})}$ \pause \color{violet} \textbf{\textsl{$\bm{\leftarrow}$ Nonparametric Additive Model}}} -->
<!-- \end{block} -->