---
title: "Sparse Multiple Index (SMI) Models for High-dimensional Nonparametric Forecasting"
author: Nuwani Palihawadana
date: 01 July 2024
titlegraphic: bg-13.png
titlecolor: white
toc: false
format:
  presentation-beamer:
    pdf-engine: xelatex
    template-partials:
        - before-title.tex
execute:
  cache: false
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| echo: false
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tsibble)
library(lubridate)
library(cowplot)
library(kableExtra)
```


# Motivation

## Daily Mortality Data
```{r}
#| label: heat-full-plot
#| echo: false
#| message: false

dataHeat <- readRDS(here::here("data/Heat_Data_Full.rds"))
dataHeat |>
  mutate(Date = as.POSIXct(Date)) |>
  ggplot(aes(x = Date, y = Death_lag_000)) +
  geom_line(colour = "#D55E00") +
  scale_x_datetime(date_breaks = "3 year") +
  labs(x = "Date", y = "Number of Deaths", title = "Daily Deaths in Montreal, Canada") +
  theme_bw()
```


## Daily Mortality Data
```{r}
#| label: heat-small-plot
#| echo: false
#| message: false

p1 <- dataHeat |>
  mutate(Date = as.POSIXct(Date)) |>
  filter(Date <= "1992-12-31") |>
  ggplot(aes(x = Date, y = Death_lag_000)) +
  geom_line(colour = "#D55E00") +
  scale_x_datetime(date_breaks = "6 months") +
  labs(x = "Date", y = "Number of Deaths", title = "Daily Deaths - 1990 to 1992") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 12, size = 7))

p2 <- dataHeat |>
  mutate(Date = as.POSIXct(Date)) |>
  filter(Date <= "1990-12-31") |>
  ggplot(aes(x = Date, y = Death_lag_000)) +
  geom_line(colour = "#D55E00") +
  scale_x_datetime(date_breaks = "2 months") +
  labs(x = "Date", y = "Number of Deaths", title = "Daily Deaths - 1990") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 12, size = 7))  +
  geom_smooth(colour = "black", se = FALSE, size = 0.4)

plot_grid(p1, p2, nrow=1)
```


## Daily Mortality Data
```{r}
#| label: temp-heat-plot
#| echo: false
#| message: false

dataSummer <- readRDS(here::here("data/Heat_Corrected.rds"))

p3 <- dataSummer |>
  ggplot(aes(x = Tmax_lag_000, y = Death_lag_000)) +
  geom_point(colour = "#CC79A7", alpha = 0.3) +
  xlim(c(0, 35)) +
  labs(
    x = "Maximum Temperature (Degrees Celsius)",
    y = "Number of Deaths",
    title = "Deaths (Summer) against Max Temperature"
  ) +
  theme_bw() +
  geom_smooth(colour = "black", se = FALSE, size = 0.4)

p4 <- dataSummer |>
  ggplot(aes(x = Tmin_lag_000, y = Death_lag_000)) +
  geom_point(colour = "#CC79A7", , alpha = 0.3) +
  xlim(c(0, 35)) +
  labs(
    x = "Minimum Temperature (Degrees Celsius)",
    y = "Number of Deaths",
    title = "Deaths (Summer) against Min Temperature"
  ) +
  theme_bw() +
  geom_smooth(colour = "black", se = FALSE, size = 0.4)

plot_grid(p3, p4, nrow=1)
```


# Background

## Background

\begin{itemize}
  \item \color{violet} \textbf{\textsl{Nonlinear "Transfer Function" model}}
\end{itemize}
\begin{block}{}
  \centerline{$y_{t} = f(\mathbfit{x}_{t}, \mathbfit{x}_{t-1}, \dots,\mathbfit{x}_{t-p}, y_{1},\dots,y_{t-k}) + \varepsilon_{t}$}
\end{block}

$y_{t}$ -- variable to forecast \newline
$\mathbfit{x}_{t}$ -- a vector of predictors \newline
$\varepsilon_{t}$ -- random error

\pause
\begin{itemize}
  \item Impossible to estimate $f$ for large $p$ -- \textbf{\textit{curse of dimensionality}}
  \pause
  \item Reasonable to impose additivity constraints \\[0.2cm]
\end{itemize}
\begin{block}{}
  \centerline{$f(\mathbfit{x}_{t}, \mathbfit{x}_{t-1}, \dots,\mathbfit{x}_{t-p}) = \sum_{i=0}^{p} {f_{i}(\mathbfit{x}_{t-i})}$ \pause \color{violet} \alert{$\mathbfit{\leftarrow}$ Nonparametric Additive Model}}
\end{block}


## Background

::: {.callout-warning}
## \color{violet} \textbf{Issues:}

\begin{enumerate}
  \item Challenging to estimate in a high-dimensional setting \newline
  \item Subjectivity in predictor selection, and predictor grouping to model interactions
\end{enumerate}
:::

\pause

::: {.callout-note}
## \color{violet} \textbf{Index Models:}

\begin{itemize}
    \item \textbf{Mitigate difficulty of estimating a nonparametric component for each predictor}
    \item \textbf{Improve flexibility}
\end{itemize}
:::


# Sparse Multiple Index (SMI) Model

## SMI Model

\begin{block}{Semi-parametric model}
$$y_{i} = \beta_{0} + \sum_{j = 1}^{p}g_{j}(\mathbfit{\alpha}_{j}^{T}\mathbfit{x}_{ij}) + \sum_{k = 1}^{d}f_{k}(w_{ik}) + \mathbfit{\theta}^{T}\mathbfit{u}_{i} + \varepsilon_{i}, \quad i = 1, \dots, n,$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{i}$ -- \color{violet} univariate response
  \item \small \color{black} $\mathbfit{x}_{ij} \in \mathbb{R}^{\ell_{j}}$, $j = 1, \dots, p$ -- \color{violet} $p$ subsets of predictors entering indices
  \item \small \color{black} $\mathbfit{\alpha}_{j}$ -- \color{violet} $\ell_{j}$-dimensional vectors of index coefficients
  \item \small \color{black} $g_{j}, f_{k}$ -- \color{violet} smooth nonlinear functions
  \item \small \color{black} Additional predictors :
    \begin{itemize}
      \item \small $w_{ik}$  --   \color{violet} nonlinear 
      \item \small \color{black} $\mathbfit{u}_{i}$  --   \color{violet} linear
    \end{itemize}
\end{itemize}


## SMI Model

\begin{block}{Semi-parametric model}
$$y_{i} = \beta_{0} + \sum_{j = 1}^{p}g_{j}(\mathbfit{\alpha}_{j}^{T}\mathbfit{x}_{ij}) + \sum_{k = 1}^{d}f_{k}(w_{ik}) + \mathbfit{\theta}^{T}\mathbfit{u}_{i} + \varepsilon_{i}, \quad i = 1, \dots, n,$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{i}$ -- \color{violet} univariate response
  \item \small \color{black} $\mathbfit{x}_{ij} \in \mathbb{R}^{\ell_{j}}$, $j = 1, \dots, p$ -- \color{violet} $p$ subsets of predictors entering indices
  \item \small \color{black} $\mathbfit{\alpha}_{j}$ -- \color{violet} $\ell_{j}$-dimensional vectors of index coefficients
  \item \small \color{black} $g_{j}, f_{k}$ -- \color{violet} smooth nonlinear functions
  \item \small \color{black} Additional predictors :
    \begin{itemize}
      \item \small $w_{ik}$  --   \color{violet} nonlinear 
      \item \small \color{black} $\mathbfit{u}_{i}$  --   \color{violet} linear
    \end{itemize}
\end{itemize}

\begin{textblock}{5.5}(9, 6)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\scriptsize Both "p" and the predictor grouping among indices are unknown.
\end{alertblock}
\end{textblock} 

\begin{textblock}{5.5}(9, 7)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\scriptsize Overlapping of predictors among indices is not allowed.
\end{alertblock}
\end{textblock} 


## Optimisation Problem

\small Let \color{violet}\textbf{$q$} \color{black}be the \color{violet}\textit{total number of predictors} \color{black}entering indices.

\begin{block}{}
\begin{align*}
  & \min_{\beta_{0}, p, \mathbfit{\alpha}, \mathbfit{g}, \mathbfit{f}, \mathbfit{\theta}} \quad \sum_{i = 1}^{n}\Bigg [ y_{i} - \beta_{0} - \sum_{j = 1}^{p}g_{j}(\mathbfit{\alpha}_{j}^{T}\mathbfit{x}_{i}) - \sum_{k = 1}^{d}f_{k}(w_{ik}) - \mathbfit{\theta}^{T}\mathbfit{u}_{i}\Bigg]^{2} \\
  & \hspace*{5cm} + \color{blue}\lambda_{0}\sum_{j = 1}^{p}\sum_{m = 1}^{q}\mathbb{1}(\alpha_{jm} \neq 0) \color{black}+ \color{magenta}\lambda_{2}\sum_{j = 1}^{p}\|\mathbfit{\alpha}_{j}\|_2^2 \nonumber \\
  & \hspace*{1.5cm} \text{s.t.}\quad \sum_{j=1}^p \mathbb{1}(\alpha_{jm} \neq 0) \in \{0,1\} \quad \forall m \nonumber
\end{align*}
\end{block}
\pause
\begin{itemize}
  \item \color{blue}$\lambda_{0} > 0$ -- controls the number of selected predictors
  \pause
  \item \color{magenta}$\lambda_{2} \ge 0$ -- controls the strength of the additional shrinkage
\end{itemize}


## MIQP Formulation

\begin{block}{}
$$
\begin{aligned}
  \min_{\beta_{0}, p, \mathbfit{\alpha}, \mathbfit{g}, \mathbfit{f}, \mathbfit{\theta}, \mathbfit{z}} \quad & \sum_{i = 1}^{n}\Bigg [ y_{i} - \beta_{0} - \sum_{j = 1}^{p}{g_{j}(\mathbfit{\alpha}_{j}^{T}\mathbfit{x}_{i})} - \sum_{k = 1}^{d} {f_{k}(w_{ik})} - \mathbfit{\theta}^{T}\mathbfit{u}_{i}\Bigg ]^{2} \\
  & \hspace*{4cm} + \color{blue}\lambda_{0}\sum_{j = 1}^{p}\sum_{m = 1}^{q}z_{jm} \color{black}+ \color{magenta}\lambda_{2}\sum_{j = 1}^{p}\sum_{m = 1}^{q} \alpha_{jm}^{2} \\
  \text{s.t.} \quad & |\alpha_{jm}| \le Mz_{jm} \quad \forall j, \forall m, \\
  & \sum_{j = 1}^{p}z_{jm} \le 1 \quad \forall m, \\
  & z_{jm} \in \{0, 1\} \pause \color{red} \quad \mathbfit{\leftarrow} \quad z_{jm} = \mathbb{1}(\alpha_{jm} \neq 0)
\end{aligned}
$$
\end{block}
\pause
\begin{itemize}
  \item $M < \infty$: \small{If $\mathbfit{\alpha^{*}}$ is an optimal solution, then $\max \big ( \{|\alpha_{jm}^{*} |\}_{j\in[p],m\in[q]} \big) \le M$}
\end{itemize}

## Estimation Algorithm

\begin{block}{}
\color{blue} \textbf{Step 1 :} \color{black} Initialise index structure and index coefficients
\end{block}
\pause
\begin{block}{}
\color{violet} \textbf{Step 2 :} \color{black} Estimate nonlinear functions
\end{block}
\pause
\begin{block}{}
\color{blue} \textbf{Step 3 :} \color{black} Update index coefficients
\end{block}
\pause
\begin{block}{}
\color{violet} \textbf{Step 4 :} \color{black} Iterate steps 2 and 3
\end{block}
\pause
\begin{block}{}
\color{blue} \textbf{Step 5 :} \color{black} Add a new index with dropped predictors, and repeat step 4
\end{block}
\pause 
\begin{block}{}
\color{violet} \textbf{Step 6 :} \color{black} Increase p by 1 in each iteration of step 5  until reaching termination criteria
\end{block}


## Forecasting Heat Exposure Related Mortality

\begin{textblock}{7}(0.7, 1)
\fontsize{11}{12}\sf
\begin{block}{Variables}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS, Year \newline
  \end{itemize}
\end{block}
\end{textblock}


# Empirical Applications

## Forecasting Heat Exposure Related Mortality

\begin{textblock}{7}(0.7, 1)
\fontsize{11}{12}\sf
\begin{block}{Variables}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS, Year \newline
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{7}(8.3, 1)
\fontsize{11}{12}\sf
\begin{block}{Data Split}
  \begin{itemize}
  \item \color{violet}\textbf{Training Set:} \color{black}1990 to 2012 \newline
  \item \color{violet}\textbf{Validation Set:} \color{black}2013 \newline
  \item \color{violet}\textbf{Test Set:} \color{black}2014 \newline \newline \newline \newline \newline
\end{itemize}
\end{block}
\end{textblock}


# Empirical Applications

## Forecasting Heat Exposure Related Mortality

\begin{textblock}{7}(0.7, 1)
\fontsize{11}{12}\sf
\begin{block}{Variables}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS, Year \newline
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{7}(8.3, 1)
\fontsize{11}{12}\sf
\begin{block}{Data Split}
  \begin{itemize}
  \item \color{violet}\textbf{Training Set:} \color{black}1990 to 2012 \newline
  \item \color{violet}\textbf{Validation Set:} \color{black}2013 \newline
  \item \color{violet}\textbf{Test Set:} \color{black}2014 \newline \newline \newline \newline \newline
\end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{14.6}(0.7, 6.5)
\fontsize{11}{12}\sf
\begin{alertblock}{}
$$
  \textbf{Deaths} = \beta_{0} + \sum_{j = 1}^{p}{g_{j}(\mathbfit{X}\mathbfit{\alpha}_{j})} + f_{1}(\textbf{DOS}) + f_{2}(\textbf{Year})+ \mathbfit{\varepsilon},
$$
\end{alertblock}
\end{textblock}

## Results

\fontsize{10}{12}\sf
```{r}
#| echo: false
results_heat <- readr::read_csv("results/heat_results_correct.csv")
kable(results_heat,
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    escape = FALSE,
    linesep = "",
    col.names = c("Model", "Predictors", "Indices", "MSE", "MAE", "MSE", "MAE")
  ) |>
  add_header_above(c("", "", "", "Test Set 1" = 2, "Test Set 2" = 2), align = "c") |>
  kable_styling(latex_options = c("repeat_header")) |>
  row_spec(0, align = "c") |>
  column_spec(4, bold = if_else(results_heat$MSE1 == min(results_heat$MSE1), TRUE, FALSE)) |>
  column_spec(5, bold = if_else(results_heat$MAE1 == min(results_heat$MAE1), TRUE, FALSE)) |>
  column_spec(6, bold = if_else(results_heat$MSE2 == min(results_heat$MSE2), TRUE, FALSE)) |>
  column_spec(7, bold = if_else(results_heat$MAE2 == min(results_heat$MAE2), TRUE, FALSE))
```

\begin{textblock}{14}(1, 7)
\fontsize{11}{12}\sf
\begin{block}{}
\begin{itemize}
  \item \color{violet} \textbf{Test Set 1:} \color{black} Three months (June, July and August 2014)
  \item \color{violet} \textbf{Test Set 2:} \color{black} One month (June 2014)
\end{itemize}
\end{block}
\end{textblock}


# Conclusion

## Conclusion

::: {.callout-note}
## \color{violet} \textbf{Key features:}

\begin{itemize}
    \item Automatic selection of number of indices and predictor grouping
    \item Automatic predictor selection
    \item A wide spectrum: from single index models to additive models
    \item Flexibility to include separate nonlinear and linear predictors
  \end{itemize}
:::

\pause

::: {.callout-warning}
## \color{violet} Things to improve:

\begin{itemize}
    \item \small Initialisation: we encourage trial-and-error
    \item \small Applicability: more applications are needed
    \item \small Computational time: increases with number of predictors and indices
  \end{itemize}
:::

## 

\placefig{1.1}{1}{width=3cm}{hex-large}

\begin{textblock}{10.2}(4.8, 1)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\begin{itemize}
  \item \color{blue} \textbf{R package :}
\end{itemize}
  \hspace*{0.8cm} \href{https://github.com/nuwani-palihawadana/smimodel}{github.com/nuwani-palihawadana/smimodel}
\end{alertblock}
\end{textblock}

\begin{textblock}{10.2}(4.8, 2.8)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\begin{itemize}
  \item \color{blue} \textbf{Paper :}
\end{itemize}
  \hspace*{0.8cm} \href{https://github.com/nuwani-palihawadana/smimodel_paper}{github.com/nuwani-palihawadana/smimodel\_paper}
\end{alertblock}
\end{textblock}

\begin{textblock}{10.2}(2.9, 5)
\fontsize{12}{12}\sf
\begin{block}{Find me :}
\faLinkedin \hspace*{0.3cm} \href{https://www.linkedin.com/in/nuwani-palihawadana/}{in/nuwani-palihawadana} \newline
\faGithub \hspace*{0.3cm} \href{https://github.com/nuwani-palihawadana}{@nuwani-palihawadana} \newline
\faEnvelope \hspace*{0.3cm} \href{mailto:nuwani.kodikarapalihawadana@monash.edu}{nuwani.kodikarapalihawadana@monash.edu}
\end{block}
\end{textblock}